Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)\\ Technical Appendix Including Proofs and Details

### 2017-08-21 fully-abstract-comp/technical-report.tex
updated call semantics
Added some description and formatted the definition a little bit better

%%% Math notation
\newcommand{\defeq}{\stackrel{\textit{\tiny{def}}}{=}}
\newcommand{\defbnf}{::=}
\newcommand{\sem}[1]{\left\llbracket #1 \right\rrbracket}
\newcommand{\ssem}[2][\Phi]{\sem{#2}_{\mathrm{src}}(#1)}

@@

\newcommand{\failed}{\mathrm{failed}}
\newcommand{\halted}{\mathrm{halted}}

@@

%%% Variables
\newcommand{\var}[1]{\mathit{#1}}
\newcommand{\rn}{\var{rn}}
\newcommand{\reg}{\var{reg}}
\newcommand{\mem}{\var{mem}}
\newcommand{\ms}{\var{ms}}
\newcommand{\pc}{\var{pc}}
\newcommand{\stk}{\var{stk}}
\newcommand{\opc}{\var{opc}}
\newcommand{\vsc}{\var{sc}}
\newcommand{\cb}{\var{cb}}

@@

%%% Functions
\newcommand{\plainfun}[2]{
  \ifthenelse{\equal{#2}{}}
  {\mathit{#1}}
  {\mathit{#1}(#2)}
}
\newcommand{\updPcAddr}[1]{\plainfun{updatePc}{#1}}
\newcommand{\updPcPerm}[1]{\plainfun{updatePcPerm}{#1}}
\newcommand{\nonexec}[1]{\plainfun{nonExecutable}{#1}}
\newcommand{\noRetStkReg}[1]{\plainfun{noRetStk_{reg}}{#1}}
\newcommand{\noRetStkMs}[1]{\plainfun{noRetStk_{ms}}{#1}}

\begin{document}
\section{The two capability machines}
\subsection{Domains}
Source language domains:
\[
  \begin{array}{rrcl}
   \addrbnf,\basebnf \in & \Addr & \defeq & \nats \\
    \sealbasebnf, \sigma \in & \Seal & \defeq & \nats \\
    &\Word & \defeq & \ints \uplus \Caps\\
    \permbnf \in& \Perm & \defbnf & \dots \\
    &\glbnf & \defbnf & \glob \mid \local \\
    & \aendbnf & \in & \Addr \uplus \{\infty \} \\
    &  \sealendbnf & \in & \Seal \uplus \{\infty \} \\
    &\scbnf &\defbnf & ((\permbnf,\glbnf),\basebnf,\aendbnf,\addrbnf) \mid \seal{\glbnf,\sealbasebnf,\sealendbnf}\\
    & & & {\sourcecolor{} \mid \stkptr{\permbnf,\basebnf,\aendbnf,\addrbnf}}\\     &\cbnf & \defbnf &  \scbnf \mid \sealed{\scbnf,\sigma}{\sourcecolor{} \; \mid \retptrd \mid \retptrc}\\ 
    &\RegName & \defbnf &\pcreg \mid \rO \mid \rstk \mid \rdata \mid \dots \\
    &\Reg & \defeq & \RegName \fun \Word\\
    &\Mem & \defeq & \Addr \fun \Word \\
    &\MemSeg & \defeq & \Addr \parfun \Word \\
    {\sourcecolor{} \sstkframe \in} & {\sourcecolor{} \StkFrame} & {\sourcecolor{} \defeq} & {\sourcecolor{} \Caps \times \MemSeg \times \Addr}\\
    {\sourcecolor{} \sstk \in}& {\sourcecolor{} \Stack} & {\sourcecolor{} \defeq} & {\sourcecolor{} \StkFrame^*} \\
    \Phi \in & \ExecConf & \defeq & \Mem \times \Reg {\sourcecolor{} \; \times \; \Stack \times \MemSeg }\\
    &\Conf & \defeq & \ExecConf \uplus \{\failed\} \uplus (\{\halted\} \times \Mem)
  \end{array}
\]
The target language domains are all the non blue parts in the above. The source language domains are the black and blue parts in the above. Further
\begin{itemize}
\item $\glbnf$ defines domain $\Global$
\item $\scbnf$ defines domain $\SealableCaps$
\item $\cbnf$ defines domain $\Caps$
\end{itemize}

@@

\subsection{Syntax}
The target machine is a simple capability machine with memory capabilities, local capabilities and sealed capabilities\footnote{In previous work, we used enter capabilities but for this the complexity introduced by mixing writable and executable memory is difficult to handle.} (Inspired by CHERI). The syntax of the instructions of the target machine is defined as follows:
\[
\begin{array}{rcl}
n &\in & \nats \\
\trg{r} &\in &  \tRegName \\
\trg{\rn} &::= &  \trg{r} \mid n \\
\trg{i} &::= & \tfail \mid \thalt \mid \tjmp{\trg{r}} \mid \tjnz{\trg{r}}{\trg{\rn}} \mid \tisptr{\trg{r}}{\trg{r}} \mid \tgeta{\trg{r}}{\trg{r}} \mid \tgetb{\trg{r}}{\trg{r}} \mid \\
      & &  \tgete{\trg{r}}{\trg{r}}\mid \tgetp{\trg{r}}{\trg{r}} \mid \tgetl{\trg{r}}{\trg{r}} \mid \tmove{\trg{r}}{\trg{\rn}} \mid \tstore{\trg{r}}{\trg{r}} \mid\\
      & &  \tload{\trg{r}}{\trg{r}} \mid \tlea{\trg{r}}{\trg{\rn}} \mid \trestrict{\trg{r}}{\trg{r}}{\trg{\rn}} \mid \tsubseg{\trg{r}}{\trg{\rn}}{\trg{\rn}} \mid \tlt{\trg{r}}{\trg{r}}{\trg{r}} \mid \\
      & & \tplus{\trg{r}}{\trg{\rn}}{\trg{\rn}} \mid \tminus{\trg{r}}{\trg{\rn}}{\trg{\rn}} \mid \tsetatob{\trg{r}} \mid \txjmp{r}{r} \mid \tcseal{r}{r}{r}
\end{array}
\]

The source machine is also a capability machine with memory capabilities, local capabilities and sealed capabilities. Unlike the target machine, the source machine is going to have a built in stack along with special stack and return capabilities. The syntax of the source machine language is as follows:
\[
  \begin{array}{rcl}
    n & \in & \nats \\
    \src{r} &\in &  \RegName \setminus \{\rO,\rstk\}\\
    \src{\rn} &::= & \src{r} \mid n \\
    \src{i} & ::= &  \trg{i} \mid \scall{\src{r}}{\src{r}}
  \end{array}
\]
There is one syntactic difference between the source language and the target language, namely the target language has an extra instruction in the $\scall{}{}$ instruction.

@@

\[
  \nonexec{} \defeq \dots
\]

\[
  \noRetStkReg{} \defeq \dots %remember to ignore certain registers
\]

\[
  \noRetStkMs{} \defeq \dots %remember to ignore certain registers
\]

\subsubsection{Source Language}
\lau{23-03-2017: $\sjmp{}$ and $\scall{}{}$ needs an update.}
\begin{align*}
  \ssem{\scall{r_1}{r_2}} = & 
    \begin{cases}
      (\mem,\reg',\stk',\ms_\stk') & \Phi = (\mem,\reg,\stk,\ms_\stk) \text{ and } \dots\\
      \\
      \failed & \text{otherwise}
    \end{cases}
\end{align*}

@@

Unlike CHERI, we want sealed capabilies to be useable in a somewhat normal jump as well as the $\scall{}{}$. We cannot just use $\sjmp{}$ because it only takes one argument and we need to be able to jump to a code/data pair. To this end we introduce $\sxjmp{}{}$:
\begin{align*}
  \ssem{\sxjmp{r_1}{r_2}} = & 
                              \begin{cases}
                                \Phi
                                \arraycolsep=0pt
                                \begin{array}[t]{l}
                                  [\reg.\pcreg \mapsto c_1] \\
                                  [\reg.\rdata \mapsto c_2]
                                \end{array} & 
                                \arraycolsep=0pt
                                \begin{array}[t]{l}
                                  \Phi(r_1) = \sealed{c_1,\sigma} \text{ and } \\
                                  \Phi(r_2) = \sealed{c_2,\sigma} \text{ and } \\
                                  \nonexec{c_2}
                                \end{array}\\
%
                                \\
%
                                \failed & \text{otherwise}
                              \end{cases}
\end{align*}
\lau{This is $\sxjmp{}{}$ for the target language - this one also needs to allow jumps to the return pointer}
Notice, $\rdata$ serves the same function as the IDC register in CHERI, i.e.\ this is where the data capability is put. In the interpretation of $\sxjmp{}{}$, we require $c_2$ to be non-executable because CHERI does it. We are, however, not certain whether this requirement is strictly necessary (maybe it is dangerous to $\sxjmp{r}{r}$ where $r$ points to the same capability? Maybe it can be use to unseal a code capability by first sealing another code capability with the seal, $\sxjmp{}{}$ with the newly sealed and the old sealed capability. Now both capabilities are unsealed). CHERI also checks whether $c_1$ is executable and in range, but we do this on every step of the operational semantics, so we do not repeat that check.

@@

\begin{align*}
  \ssem{\sxjmp{r_1}{r_2}} = & 
                              \begin{cases}
                                \Phi
                                \arraycolsep=0pt
                                \begin{array}[t]{l}
                                  [\reg.\pcreg \mapsto c_1] \\
                                  [\reg.\rdata \mapsto c_2]
                                \end{array} & 
                                \arraycolsep=0pt
                                \begin{array}[t]{l}
                                  \Phi(r_1) = \sealed{c_1,\sigma} \text{ and } \\
                                  \Phi(r_2) = \sealed{c_2,\sigma} \text{ and } \\
                                  \nonexec{c_2}
                                \end{array}
\\
\failed & \text{otherwise}
                              \end{cases}
\end{align*}

>>>>

%%% Math notation
\newcommand{\defeq}{\stackrel{\textit{\tiny{def}}}{=}}
\newcommand{\defbnf}{::=}
\newcommand{\sem}[1]{\left\llbracket #1 \right\rrbracket}
\newcommand{\ssem}[2][\Phi]{\sem{#2}_{\mathrm{src}}(#1)}
\newcommand{\dom}{\mathrm{dom}}

@@

\newcommand{\failed}{\mathrm{failed}}
\newcommand{\undefined}{\mathrm{undefined}}
\newcommand{\halted}{\mathrm{halted}}

@@

%%% Variables
\newcommand{\var}[1]{\mathit{#1}}
\newcommand{\rn}{\var{rn}}
\newcommand{\reg}{\var{reg}}
\newcommand{\mem}{\var{mem}}
\newcommand{\ms}{\var{ms}}
\newcommand{\pc}{\var{pc}}
\newcommand{\stk}{\var{stk}}
\newcommand{\ret}{\var{ret}}
\newcommand{\data}{\var{data}}
\newcommand{\priv}{\var{priv}}
\newcommand{\opc}{\var{opc}}
\newcommand{\vsc}{\var{sc}}
\newcommand{\cb}{\var{cb}}
\newcommand{\baddr}{\var{b}}
\newcommand{\eaddr}{\var{e}}
\newcommand{\aaddr}{\var{a}}
\newcommand{\stdrng}{[\baddr,\eaddr]}

%%% Constants
\newcommand{\constant}[1]{\mathrm{#1}}
\newcommand{\calllen}{\constant{call\_len}}

@@

%%% Functions
\newcommand{\plainfun}[2]{
  \ifthenelse{\equal{#2}{}}
  {\mathit{#1}}
  {\mathit{#1}(#2)}
}
\newcommand{\updPcAddr}[1]{\plainfun{updatePc}{#1}}
\newcommand{\updPcPerm}[1]{\plainfun{updatePcPerm}{#1}}
\newcommand{\nonExec}[1]{\plainfun{nonExecutable}{#1}}
\newcommand{\nonLoc}[1]{\plainfun{nonLocal}{#1}}
\newcommand{\noRetStkReg}[1]{\plainfun{noRetStk_{reg}}{#1}}
\newcommand{\noRetStkMs}[1]{\plainfun{noRetStk_{ms}}{#1}}

\begin{document}
\section{The two capability machines}
\subsection{Domains}
Source language domains:
\[
  \begin{array}{rrcl}
   \addrbnf,\basebnf \in & \Addr & \defeq & \nats \\
    \sealbasebnf, \sigma \in & \Seal & \defeq & \nats \\
    &\Word & \defeq & \ints \uplus \Caps\\
    \permbnf \in& \Perm & \defbnf & \dots \\
    &\glbnf & \defbnf & \glob \mid \local \\
    & \aendbnf & \in & \Addr \uplus \{\infty \} \\
    &  \sealendbnf & \in & \Seal \uplus \{\infty \} \\
    &\scbnf &\defbnf & ((\permbnf,\glbnf),\basebnf,\aendbnf,\addrbnf) \mid \seal{\glbnf,\sealbasebnf,\sealendbnf}\\
    & & & {\sourcecolor{} \mid \stkptr{\permbnf,\basebnf,\aendbnf,\addrbnf}}\\     &\cbnf & \defbnf &  \scbnf \mid \sealed{\scbnf,\sigma}{\sourcecolor{} \; \mid \retptr}\\ 
    &\RegName & \defbnf &\pcreg \mid \rO \mid \rstk \mid \rdata \mid \dots \\
    &\Reg & \defeq & \RegName \fun \Word\\
    &\Mem & \defeq & \Addr \fun \Word \\
    &\MemSeg & \defeq & \Addr \parfun \Word \\
    {\sourcecolor{} \sstkframe \in} & {\sourcecolor{} \StkFrame} & {\sourcecolor{} \defeq} & {\sourcecolor{} \Caps \times \MemSeg \times \Addr}\\
    {\sourcecolor{} \sstk \in}& {\sourcecolor{} \Stack} & {\sourcecolor{} \defeq} & {\sourcecolor{} \StkFrame^*} \\
    \Phi \in & \ExecConf & \defeq & \Mem \times \Reg {\sourcecolor{} \; \times \; \Stack \times \MemSeg }\\
    &\Conf & \defeq & \ExecConf \uplus \{\failed\} \uplus (\{\halted\} \times \Mem)
  \end{array}
\]
The target language domains are all the non blue parts in the above. The source language domains are the black and blue parts in the above. Further
\begin{itemize}
\item $\glbnf$ defines domain $\Global$
\item $\scbnf$ defines domain $\SealableCaps$
\item $\cbnf$ defines domain $\Caps$
\end{itemize}

@@

\subsection{Syntax}
The target machine is a simple capability machine with memory capabilities, local capabilities and sealed capabilities\footnote{In previous work, we used enter capabilities but for this the complexity introduced by mixing writable and executable memory is difficult to handle.} (Inspired by CHERI). The syntax of the instructions of the target machine is defined as follows:
\[
\begin{array}{rcl}
n &\in & \nats \\
\trg{r} &\in &  \tRegName \\
\trg{\rn} &\defbnf &  \trg{r} \mid n \\
\trg{i} &\defbnf & \tfail \mid \thalt \mid \tjmp{\trg{r}} \mid \tjnz{\trg{r}}{\trg{\rn}} \mid \tisptr{\trg{r}}{\trg{r}} \mid \tgeta{\trg{r}}{\trg{r}} \mid \tgetb{\trg{r}}{\trg{r}} \mid \\
      & &  \tgete{\trg{r}}{\trg{r}}\mid \tgetp{\trg{r}}{\trg{r}} \mid \tgetl{\trg{r}}{\trg{r}} \mid \tmove{\trg{r}}{\trg{\rn}} \mid \tstore{\trg{r}}{\trg{r}} \mid\\
      & &  \tload{\trg{r}}{\trg{r}} \mid \tlea{\trg{r}}{\trg{\rn}} \mid \trestrict{\trg{r}}{\trg{r}}{\trg{\rn}} \mid \tsubseg{\trg{r}}{\trg{\rn}}{\trg{\rn}} \mid \tlt{\trg{r}}{\trg{r}}{\trg{r}} \mid \\
      & & \tplus{\trg{r}}{\trg{\rn}}{\trg{\rn}} \mid \tminus{\trg{r}}{\trg{\rn}}{\trg{\rn}} \mid \tsetatob{\trg{r}} \mid \txjmp{r}{r} \mid \tcseal{r}{r}{r}
\end{array}
\]

The source machine is also a capability machine with memory capabilities, local capabilities and sealed capabilities. Unlike the target machine, the source machine is going to have a built in stack along with special stack and return capabilities. The syntax of the source machine language is as follows:
\[
  \begin{array}{rcl}
    n & \in & \nats \\
    \src{r} &\in &  \RegName \setminus \{\rO,\rstk\}\\
    \src{\rn} &\defbnf & \src{r} \mid n \\
    \src{i} & \defbnf &  \trg{i} \mid \scall{\src{r}}{\src{r}}
  \end{array}
\]
There is one syntactic difference between the source language and the target language, namely the target language has an extra instruction in the $\scall{}{}$ instruction.

@@

\[
  \nonExec{\cb} \defeq \dots
\]

\[
  \noRetStkReg{\reg} \defeq \forall r \in \reg \setminus \{ \pcreg, \rstk, \rO, \rdata \} \ldotp \reg(r) \neq \stkptr(\_)
\]

\[
  \noRetStkMs{\ms} \defeq \defeq \forall \aaddr \in \dom(\ms) \ldotp \ms(\aaddr) \neq \stkptr(\_)
\]

\subsubsection{Source Language}
\lau{23-03-2017: $\sjmp{}$ and $\scall{}{}$ needs an update.}
\textbf{Call}\\
Call has three cases, a successful call, undefined and failed. For a call to be successful, it needs two sealed capabilities - a data and a code capability. The code capability must be non-executable and the stack register must contain a stack pointer. Futher the code and data capability must be local and there must be no attempt at passing on the stack capability.
\lau{Add more description here.}

For the call to be undefined, the conditions necessary for a legal call must be met and there must be an attempt at passing the stack pointer to the callee or the called capabilities must be non-local \lau{I have forgotten why this is an issue, but I think it is because we are crossing a security domain?}

If we are not in one of the two cases above, then the execution fails.
\begin{align*}
  \ssem{\scall{r_1}{r_2}} = & 
    \begin{cases}
      (\mem,\reg',\stk',\ms_\stk') & \arraycolsep=0pt
                                      \begin{array}[t]{l}
                                        % "red conditions" (reference to blackboard picture)
                                        \Phi = (\mem,\reg,\stk,\ms_\stk) \text{ and } \\
                                        \Phi(r_1) = \sealed{\cb_1,\sigma} \text{ and } \\
                                        \Phi(r_2) = \sealed{\cb_2,\sigma} \text{ and } \\
                                        \nonExec{cb_2} \text{ and } \\
                                        \Phi(\rstk) = \stkptr{\perm, \stdrng, \aaddr} \text{ and } \\
                                        % "yellow conditions"
                                        \cb_2 \neq \stkptr{\_,\_,\_,\_} \text{ and } \\
                                        \noRetStkReg{\reg} \text{ and } \\
                                        \noRetStkMs{\ms_\stk'} \text{ and } \\
                                        \nonLoc{\cb_1,\cb_2} \text{ and } \\
                                        \reg' = \reg
                                        \begin{array}[t]{l}
                                        \update{\pcreg}{\cb_1}\update{\rstk}{\stkptr{\perm,[\aaddr,eaddr],\aaddr}}\\
                                        \update{\rO}{\retptr}\update{\rdata}{\cb_2} \text{ and }
                                        \end{array}
 \\
                                        \opc = \Phi(\pcreg)\update{a}{\Phi(\pcreg).a+\calllen} \text{ and } \\
                                        \ms_\priv = \ms_\stk \setminus [\aaddr,\eaddr] \text{ and } \\
                                        \stk' = (\opc,\ms_\priv,\perm,\stdrng,\aaddr) :: \stk \text{ and } \\
                                        \ms_\stk' = \ms_\stk |_{[\baddr,\eaddr]}
                                      \end{array}\\
                                      & \\
      \undefined & \arraycolsep=0pt
                                      \begin{array}[t]{l}
                                        % "red conditions" (reference to blackboard picture)
                                        \Phi = (\mem,\reg,\stk,\ms_\stk) \text{ and } \\
                                        \Phi(r_1) = \sealed{\cb_1,\sigma} \text{ and } \\
                                        \Phi(r_2) = \sealed{\cb_2,\sigma} \text{ and } \\
                                        \nonExec{cb_2} \text{ and } \\
                                        \Phi(\rstk) = \stkptr{\perm, \stdrng, \aaddr} \text{ and } \\
                                        % "yellow conditions"
                                        \text{not } \arraycolsep=0pt
                                        \left(\begin{array}{l}
                                          \cb_2 \neq \stkptr{\_,\_,\_,\_} \text{ and } \\
                                          \noRetStkReg{\reg} \text{ and } \\
                                          \noRetStkMs{\ms_\stk'} \text{ and } \\
                                          \nonLoc{\cb_1,\cb_2}
                                        \end{array} \right)
                                      \end{array}\\
                                      & \\
      \failed & \text{otherwise}
    \end{cases}
\end{align*}

@@

Unlike CHERI, we want sealed capabilies to be useable in a somewhat normal jump as well as the $\scall{}{}$. We cannot just use $\sjmp{}$ because it only takes one argument and we need to be able to jump to a code/data pair. To this end we introduce $\sxjmp{}{}$:
\begin{align*}
  \ssem{\sxjmp{r_1}{r_2}} = & 
                              \begin{cases}
                                \Phi
                                \arraycolsep=0pt
                                \begin{array}[t]{l}
                                  [\reg.\pcreg \mapsto c_1] \\
                                  [\reg.\rdata \mapsto c_2]
                                \end{array} & 
                                \arraycolsep=0pt
                                \begin{array}[t]{l}
                                  \Phi(r_1) = \sealed{c_1,\sigma} \text{ and } \\
                                  \Phi(r_2) = \sealed{c_2,\sigma} \text{ and } \\
                                  \nonExec{c_2}
                                \end{array}\\
%
                                \\
%
                                \failed & \text{otherwise}
                              \end{cases}
\end{align*}
\lau{This is $\sxjmp{}{}$ for the target language - this one also needs to allow jumps to the return pointer}
Notice, $\rdata$ serves the same function as the IDC register in CHERI, i.e.\ this is where the data capability is put. In the interpretation of $\sxjmp{}{}$, we require $c_2$ to be non-executable because CHERI does it. We are, however, not certain whether this requirement is strictly necessary (maybe it is dangerous to $\sxjmp{r}{r}$ where $r$ points to the same capability? Maybe it can be use to unseal a code capability by first sealing another code capability with the seal, $\sxjmp{}{}$ with the newly sealed and the old sealed capability. Now both capabilities are unsealed). CHERI also checks whether $c_1$ is executable and in range, but we do this on every step of the operational semantics, so we do not repeat that check.

@@

\begin{align*}
  \ssem{\sxjmp{r_1}{r_2}} = & 
                              \begin{cases}
                                \Phi
                                \arraycolsep=0pt
                                \begin{array}[t]{l}
                                  [\reg.\pcreg \mapsto c_1] \\
                                  [\reg.\rdata \mapsto c_2]
                                \end{array} & 
                                \arraycolsep=0pt
                                \begin{array}[t]{l}
                                  \Phi(r_1) = \sealed{c_1,\sigma} \text{ and } \\
                                  \Phi(r_2) = \sealed{c_2,\sigma} \text{ and } \\
                                  \nonExec{c_2}
                                \end{array}
\\
\failed & \text{otherwise}
                              \end{cases}
\end{align*}
